---
layout: page
title: Practical 8
permalink: /practical8/
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "",
                      comment="", 
                      prompt=TRUE)
Sys.setlocale("LC_TIME", "C")
```

# Objectives

The learning objectives for this practical are:

  * How to create and use list objects.
  * How to perform implicit looping through lists.
  * Add new columns to data frames.
  * How to merge data frames.
  * Learn to explore and visualize data in different ways.

# Setup and background

To do this practical you need an installation of R and RStudio. You can find
the instructions in the [setup](/setup#r-and-rstudio) link on how to install R
and RStudio in your system. For a smooth development of this practical, it is
strongly recommended that you follow and finish the previous
[practical 7](/practical7/).

We will download COVID19 vaccination and demographic data for Catalonia to
illustrate some data wrangling in R and RStudio. Please follow the next two
steps:

* __COVID19 vaccination by municipality__:

   1. Go to the Catalan Health Departament COVID19 data portal at
      [https://dadescovid.cat](https://dadescovid.cat) and switch the language to
      "ENGLISH" using the pull-down menu on the top-right corner of the page.
   2. Follow the downloads link and on the next page, go to the section 
      "Other Downloads" and click and download the "CSV" file corresponding to
      "Vaccination for COVID-19: administered doses by municipality".
      Make sure you know exactly where in your filesystem this file
      has been downloaded. **Tip:** some browsers automatically download files
      into a folder called "Downloads" or under a name corresponding to the
      translation of "Downloads" to the default language of your operating system.
   3. Make a directory in your filesystem, for instance at your _home_ directory,
      called `practical8` and copy in it the downloaded file.
   4. **Change the name of the file you just downloaded to `dosis_municipi.csv`**, 
      so that you finally have a file called `dosis_municipi.csv` in the 
      directory `practical8`.
      
* __Population by municipality__:

   1. Download the Catalan Urbanistic Map dataset at the
      [Dades Obertes de Catalunya](https://governobert.gencat.cat/ca/dades_obertes)
      data portal. You should follow the following
      [link](https://analisi.transparenciacatalunya.cat/en/Urbanisme-infraestructures/Dades-del-mapa-urban-stic-de-Catalunya/epsm-zskb) 
      and donwload the data by going to the top right corner and select "Export"
      and then "CSV". This dataset includes information of the population of Catalonia
      by municipality.
   2. Copy the donwloaded file to the `practical8` directory and
      **change its name to `poblacio_municipis.csv`**.

If you are using the UPF [myapps](https://myapps.upf.edu) cloud to run RStudio,
then you need to either use an internet browser in _myapps_ to download the data
file directly in the _myapps_ cloud or upload to the _myapps_ cloud the file
that you have downloaded in your own computer.

```{r, include=FALSE, eval=FALSE}
## Download required files
download.file("https://analisi.transparenciacatalunya.cat/api/views/irki-p3c7/rows.csv?accessType=DOWNLOAD&bom=true&format=true&delimiter=%3B&sorting=true", destfile="dosis_municipi.csv")
download.file("https://analisi.transparenciacatalunya.cat/api/views/epsm-zskb/rows.csv?accessType=DOWNLOAD&sorting=true", "poblacio_municipis.csv")
```

# Reading and filtering data

Let's load the CSV file `poblacio_municipis.csv`, which contains some demographic indicators
for the 948 municipalities in Catalonia such as the population (column `Poblacio_padro`).

```{r}
pop <- read.csv("poblacio_municipis.csv")
dim(pop)
head(pop, n=3)
table(pop$Any)
```

We can observe that this dataset contains data from different years. To continue with our
analysis, we will select the most recent data corresponding to 2020 and, moreover, we will
only keep the columns `Codi_ine_5_txt` (identifier for the municipality), `NomMun`, `Comarca`, 
`Poblacio_padro` and `Superficie_ha`.

```{r}
## build a logical mask to select for year 2020
mask <- pop$Any == 2020
## build a character string vector of the selected columns
selcols <- c("Codi_ine_5_txt", "NomMun", "Comarca", "Poblacio_padro", "Superficie_ha")
## subset the data.frame object 'pop' for rows in 'mask' and columns in 'selcols'
pop_sel <- pop[mask, selcols]
dim(pop_sel)
head(pop_sel)
```

# Lists and implicit looping

Lists allow one to group values through their elements. Let’s say we want to
group the population number by county. We can do that using the function `split()`
to which we should give a first argument of the values we want to group and a
second argument with the grouping factor.

```{r}
pbyc <- split(pop_sel$Poblacio_padro, pop_sel$Comarca)
class(pbyc)
length(pbyc)
names(pbyc)
head(pbyc)
```

Grouping values can be useful in data analysis when we want to examine the data
separately by groups. Let’s say we want to visualize the distribution of the
population for the two counties `Terra Alta` and `Urgell`, next to each
other. We can use the function `hist()` for that purpose, creating a grid of two
plotting panes using the `par()` function, as follows:

```{r comptaltaurgell, echo=TRUE}
par(mfrow=c(1, 2))
hist(pbyc[["Terra Alta"]], xlab="Population (inhabitants)", main="Terra Alta")
hist(pbyc$Urgell, xlab="Population (inhabitants)", main="Urgell")
```
Note that in the previous code we are using the double-bracket operator `[[`
instead of the dollar `$` to access the element `Terra Alta` because this element
has an space character in its name. Try to interpret the previous plots, how many
municipalities in _Urgell_ and _Terra Alta_ have less than 5,000 inhabitants?

Now, let's calculate the mean municipality population for the county of _Terra Alta_.
Having built the previous list object, we can make that calculation applying the
function `mean()` to the corresponding element of the list:

```{r}
mean(pbyc[["Terra Alta"]])
```

Now, let's compare it with the mean municipality population for _Urgell_:

```{r}
mean(pbyc$Urgell)
```

It would be tedious to do that calculation for each different county by
writing one such function call for each element of the list. As an alternative,
we could use a `while` or `for` loop that would iterate over the elements of
the list. However, R provides a more compact way to iterating over lists,
and other objects, by using functions for _implicit_
[looping](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Looping)
such as `lapply()` or `sapply()`. These functions take a list as a first argument,
iterate through each element of that list, and at each iteration apply the function
given in the second argument. Additional arguments can be given and will be passed
to the _applied_ function.

The function `lapply()` returns again the input list with its elements replaced
by the result given by the function on each corresponding element, while the
function `sapply()` attempts to simplify the resulting data structure in that if
each element of the resulting list has length 1, then it return an atomic vector.

We can calculate the mean municipality population per county with the following
call to the `sapply()` function:

```{r}
sapply(pbyc, mean)
```

**Exercise:** calculate the total population per county and the total population
in Catalonia.

# Sorting and ordering

We have two functions in R that allow us to rearrange values in particular order:

* `sort()` returns the ordered values.
* `order()` returns a permutation which rearranges its first argument into
  ascending or descending order.

By default, these functions return an ascending order, but by setting the argument
`decreasing=TRUE`, we can obtain a descending order.

**Exercise:** Using the previous functions `sort()` and `order()` and the
columns `NomMun` (municipality name) and `Poblacio_padro` (municipality
population), find out how many inhabitants have the three most and three least
populated municipalities and their names.

# Adding new columns

In some cases we might be interested in deriving new data columns from the
existing ones. For instance, let's say we want to add a new column to the
previous `data.frame` object `pop_sel` that stores the population density of
each municipality in inhabitants per
[squared kilometer (Km2)](https://en.wikipedia.org/wiki/Square_kilometre). The
column `Superficie_ha` contains the area occupied by the municipality in
[hectares](https://en.wikipedia.org/wiki/Hectare) (ha), let's convert it first to Km2:

```{r}
km2 <- pop_sel$Superficie_ha / 100
```

Finally, let's calculate the population density dividing the number of
inhabitants by the area occupied by the municipality in Km2 and add it
to `pop_sel` as a new column called `density`:

```{r}
pop_sel$Density <- pop_sel$Poblacio_padro / km2
head(pop_sel)
```

# Combining data

One of the most common operations required to answer a question with data
is to combine two datasets in some way. Let's say we want to compare
municipalities in terms of how many vaccine does per 100,000 inhabitants
have been administered. For that purpose, we load a second dataset
corresponding to the administered COVID19 vaccine doses by municipality
in Catalonia (`dosis_municipi.csv`) as follows.

```{r}
vac <- read.csv("dosis_municipi.csv", sep=";", stringsAsFactors=FALSE)
dim(vac)
head(vac)
```
The column `FABRICANT` contains the vaccine manufacturer. Let's tally the
number of administered doses per manufacturer.

```{r}
table(vac$FABRICANT)
```
The value `No administrada` corresponds to non-administered vaccine doses.
Let's discard those and work with the corresponding subset of the data in
a `data.frame` object called `vac_admin`.

```{r}
mask <- vac$FABRICANT !="No administrada"
vac_admin <- vac[mask, ]
dim(vac_admin)
```
We want to combine the filtered vaccination data with the population data.

```{r}
dim(vac_admin)
colnames(vac_admin)

dim(pop_sel)
colnames(pop_sel)
```

Note that both datasets have different dimensions and different column names,
so we need combine them using some column that is common in both datasets.
Because our purpose is to compare vaccination rates among municipalities, we
should expect that the name of the municipality could be use to combine both
datasets.

```{r}
head(vac_admin$MUNICIPI)
head(pop_sel$NomMun)
```
However, in the first dataset the name of the municipalities is all in
uppercase, while in the second is a combination of upper and lower cases. As R
compares characters in a case-sensitive manner, we won't be able to use directly
these values to combine the data frames as we can see from the following attempt
to match names.

```{r}
mt <- match(vac_admin$MUNICIPI, pop_sel$NomMun)
head(mt, n=20)
```
However, both datasets also have a column with a municipality code:

```{r}
head(vac_admin$MUNICIPI_CODI)
head(pop_sel$Codi_ine_5_txt)

mt <- match(vac_admin$MUNICIPI_CODI, pop_sel$Codi_ine_5_txt)
head(mt, n=20)
```
Once we know what column in each dataset can be used to combine them,
we can use the `merge()` function, which by default will return only rows with
common values in those two columns. By default, this function will combine data
using columns with identical names. Since in our datasets the two columns with
the common information are named differently, we need to specify which are these
columns in each `data.frame` object by using the arguments `by.x` and `by.y`.

```{r}
vac_pop_merge <- merge(vac_admin, pop_sel, 
                       by.x = "MUNICIPI_CODI", by.y = "Codi_ine_5_txt")
dim(vac_pop_merge)
head(vac_pop_merge)
```

Now the number of rows of the output data frame `vac_pop_merge` is slightly
smaller than `vac_admin`, because it doesn't keep rows that didn't find
a match in `pop_sel`.

**Exercise:** Using `vac_pop_merge`, add a new column named `doses_100K_h` 
containing how many doses were administered each day per 100,000 inhabitants. 
The column `RECOMPTE` contains the number of administered doses.
Which municipality administered the most and the least doses and at which date?

```{r, include=FALSE}
vac_pop_merge$doses_100K_h <- vac_pop_merge$RECOMPTE / vac_pop_merge$Poblacio_padro * 100000
vac_pop_merge[order(vac_pop_merge$doses_100K_h, decreasing = TRUE),]
```

To continue with our analysis, we need to add a column containing a month as 
a factor. Thus, we repeat the steps explained in the previous [practical 7](/practical7/)
to convert the column `DATA` to a date, extract the months and convert them to 
a factor with ordered levels:

```{r}
vac_pop_merge$month <- months(as.Date(vac_pop_merge$DATA), abbreviate=TRUE)
vac_pop_merge$month <- factor(vac_pop_merge$month, levels=c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                         "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```

Next, we will select the municipalities with high population (more than 200,000
residents) and save them into a data frame called `vac_high`. Now we want to 
obtain the total doses administered per inhabitant every month in this specific
municipalities. Again, we take advantege of the combination of `split` and 
`sapply` to do this calculation. Finally, we create a new data frame with the
summarised data, including a column called `muni_type` that identifies the 
type of municipalities ("High" as in "High population") used for extracting
this values.

```{r barplotDosesHighPop}
vac_high <- vac_pop_merge[vac_pop_merge$Poblacio_padro > 200000,]
dosesh_high <- split(vac_high$doses_100K_h, vac_high$month)

total_dosesh_high <- sapply(dosesh_high, sum)

df_high <- data.frame("month"=names(total_dosesh_high),
                      "doses_h"=total_dosesh_high,
                      "muni_type"="High")

barplot(df_high$doses_h, names.arg = df_high$month,
        main="Highly populated municipalities")
```


**Exercise**: Create a data frame named `df_low` that contains the total doses
per 100,000 inhabitants administered per month in municipalities with population
smaller than 1,000 inhabitants. Make a bar plot showing the administered doses per
inhabitant per month, as we did above.

```{r barplotDosesLowPop, include = FALSE}
vac_low <- vac_pop_merge[vac_pop_merge$Poblacio_padro < 1e3,]
dosesh_low <- split(vac_low$doses_100K_h, vac_low$month)

total_dosesh_low <- sapply(dosesh_low, sum)

df_low <- data.frame("month"=names(total_dosesh_low),
                      "doses_h"=total_dosesh_low,
                      "muni_type"="Low")

barplot(df_low$doses_h, names.arg = df_low$month,
        main="Lowly populated municipalities")
```

**Exercise**: Combine the data frames `df_high` and `df_low`
into a single data frame named `df_months` (**Hint**: Make sure that `df_high`
and `df_low` have the same columns, with the same name and in the same order).
Then make a box plot showing the distribution of doses per inhabitant grouped by
the type of municipality (low or high). Do you see any difference?

```{r boxplotDosesHighLowPop, include=FALSE}
df_months <- rbind(df_high, df_low)

boxplot(df_months$doses_h ~ df_months$muni_type,
        xlab="Municipality Type (~Population)", ylab="Doses per 100,000 residents")
```
