---
layout: page
title: Practical 8
permalink: /practical8/
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "",
                      comment="", 
                      prompt=TRUE)
Sys.setlocale("LC_TIME", "C")
```

# Objectives

The learning objectives for this practical are:

  * How to create and use list objects.
  * How to perform implicit looping through lists.
  * How to merge data frames.
  * Learn to visualize data in different ways.

# Setup and background

To do this practical you need an installation of R and RStudio. You can find
the instructions in the [setup](/setup#r-and-rstudio) link on how to install R
and RStudio in your system. For a smooth development of this practical, it is
strongly recommended that you follow and finish the previous
[practical 7](/practical7/).

We will download some COVID19 vaccination data, together with Catalunya population
data to illustrate data wrangling in R and RStudio.
Please follow the next two steps:

* __COVID19 vaccination by municipality__:

   1. Go to the Catalan Health Departament COVID19 data portal at
      [https://dadescovid.cat](https://dadescovid.cat) and switch the language to
      "ENGLISH" using the pull-down menu on the top-right corner of the page.
   2. Follow the downloads link and on the next page, go to the section 
      "Other Downloads" and click and download the "CSV" file corresponding to
      "Vaccination for COVID-19: administered doses by municipality".
      Make sure you know exactly where in your filesystem this file
      has been downloaded. **Tip:** some browsers automatically download files
      into a folder called "Downloads" or under a name corresponding to the
      translation of "Downloads" to the default language of your operating system.
   3. Make a directory in your filesystem, for instance at your _home_ directory,
      called `practical8` and copy in it the downloaded file.
   4. Change the name of the file you just downloaded to `dosis_municipi.csv`, 
      so that you finally have a file called `dosis_municipi.csv` in the 
      directory `practical8`.
      
* __Population by municipality__:

   1. Download the Catalan Urbanistic Map dataset at the [Dades Obertes de Catalunya](https://governobert.gencat.cat/ca/dades_obertes)
      data portal. You should follow the following [link](https://analisi.transparenciacatalunya.cat/en/Urbanisme-infraestructures/Dades-del-mapa-urban-stic-de-Catalunya/epsm-zskb) 
      and donwload the data by going to the top right corner and select "Export" 
      and then "CSV". This dataset includes information of the population of Catalunya
      by municipality.
   2. Copy the donwloaded file to the `practical8` directory and change its name
      to `poblacio_municipis.csv`.

If you are using the UPF [myapps](https://myapps.upf.edu) cloud to run RStudio,
then you need to either use an internet browser in _myapps_ to download the data
file directly in the _myapps_ cloud or upload to the _myapps_ cloud the file
that you have downloaded in your own computer.

```{r, include=FALSE}
## Download required file
download.file("https://analisi.transparenciacatalunya.cat/api/views/irki-p3c7/rows.csv?accessType=DOWNLOAD&bom=true&format=true&delimiter=%3B&sorting=true", destfile="dosis_municipi.csv")
```


# Lists and implicit looping

We first load the vaccination dataset `dosis_municipi.csv` by using the function 
`read.csv()`. As you can see, this data set contains inormation on the administered
doses of the different COVID19, by sex and municipality.

```{r}
vac <- read.csv("dosis_municipi.csv", sep=";", stringsAsFactors=FALSE)
head(vac)

table(vac$FABRICANT)
```

Now, remove the rows that correspond to non-administered vaccines and 
save it in a data frame called `vac_admin`.

```{r}
vac_admin <- vac[vac$FABRICANT!="No administrada",]
```


Lists allow one to group values through their elements. Let's say we want
to group the number of administered doses in the previous data, by the
type (or fabricant) of the vaccine. We can do that using the function
`split()` to which we should give a first argument of the values we want
to group and a second argument with the grouping factor.

```{r}
dosesbyfabricant <- split(vac_admin$RECOMPTE, vac_admin$FABRICANT)
class(dosesbyfabricant)
length(dosesbyfabricant)
names(dosesbyfabricant)
head(dosesbyfabricant, n=3)
```

Grouping values can be useful in data analysis when we want to examine the
data separately by groups. Let's say we want to visualize the distribution
of administered doses for Pfizer and Moderna, next to each other. We can 
use the function `hist()` for that purpose, creating a grid of two 
plotting panes using the `par()` function, as follows:

```{r dosesPfizerModerna}
par(mfrow=c(1, 2))
hist(dosesbyfabricant$`BioNTech / Pfizer`, xlab="Administred Doses", main="Pfizer")
hist(dosesbyfabricant$`Moderna / Lonza`, xlab="Administred Doses", main="Moderna")
```

Note that as the name of the list elements contain spaces and the character `/` we need
to add backquote to delimit the name. We could also use double square brackets (`[[]]`)
to access a specific element in the list, for example `dosesbyfabricant[["BioNTech / Pfizer"]]`.

Now, let's calculate the mean doses administered of Moderna vaccines.
Having built the previous `list` object, we can make that calculation applying
the function `mean()` to the corresponding element of the list:

```{r}
mean(dosesbyfabricant$`Moderna / Lonza`)
```
Let's say we want to compare this value with the mean value for the Pfizer
vaccines:

```{r}
mean(dosesbyfabricant$`BioNTech / Pfizer`)
```

It would be tedious to do that calculation for each different fabricant by
writing one such function call for each element of the list. As an alternative,
we could use a `while` or `for` loop that would iterate over the elements of
the list. However, R provides a more compact way to iterating over lists,
and other objects, by using functions for _implicit_
[looping](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Looping)
such as `lapply()` or `sapply()`. These functions take a list as a first argument,
iterate through each element of that list, and at each iteration apply the function
given in the second argument. Additional arguments can be given and will be passed
to the _applied_ function.

The function `lapply()` returns again the input list with its elements replaced
by the result given by the function on each corresponding element, while the
function `sapply()` attempts to simplify the resulting data structure in that if
each element of the resulting list has length 1, then it return an atomic vector.

We can calculate the mean of the risk of outbreak per month with the following
call to the `sapply()` function:

```{r}
sapply(dosesbyfabricant, mean, na.rm=TRUE)
```
where here the argument `na.rm=TRUE` is passed by `sapply()` to each call to
the `mean()` function. This argument excludes `NA` (missing) values and do the
calculation with the non-missing ones. Try the same call using the `lapply()` function 
and notice the difference in the output.

**Exercise:**

Plot the distribution of the number of administered Pfizer vaccines in 
Barcelona (municipality) per month, using box plots (**hint**: Check the materials
on [practical 7](https://funcompbio.github.io/practical7/) to remember how to 
extract months from dates and convert them to ordered factors).

Calculate the mean of the number of administered Pfizer doses per month in 
Barcelona and plot it over the previous box plots using the function `points()`.
Check the help page of `points()` to find out how to use it, figure out how 
make the plotted point to be a red solid diamond
(**hint:** look at the argument `pch`).

Your plot should look like the one below:

```{r pfizerBCNperMonth, echo = FALSE}
vac_bcn_pfi <- vac_admin[vac_admin$MUNICIPI == "BARCELONA" & vac_admin$FABRICANT == "BioNTech / Pfizer",]

## Select month info
m <- months(as.Date(vac_bcn_pfi$DATA), abbreviate = TRUE)
mf <- factor(m, levels=c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                         "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))


vac_bcn_mf <- split(vac_bcn_pfi$RECOMPTE, mf)
mean <- sapply(vac_bcn_mf, mean)

boxplot(vac_bcn_pfi$RECOMPTE ~ mf,
        xlab = "Month", ylab="Administered Vaccines",
        main = "Administered Pfizer Vaccines in Barcelona")
points(mean, pch=18, col="red", cex=2)
```

# Data wrangling

